Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-06-06 _tex/paper2.tex
Built site for gh-pages

To address these issues, this study extends Thurstone's general form
through a systematic and integrated approach that combines causal and
Bayesian inference methods. In addition to potentially enhancing
measurement reliability and validity, and improving statistical accuracy
in hypothesis testing, this approach offers two key advantages. First,
it clarifies the interactions among all actors and processes involved in
CJ assessments. Second, it shifts the current comparative data analysis
paradigm from passively accepting the BTL model assumptions to actively
testing whether those assumptions fit the data under analysis.

@@

Significant statistical and measurement issues can arise when the
assumption of equal dispersions between stimuli does not hold.
Specifically, the BTL model may overestimate the trait's reliability,
that is, the degree to which the outcome accurately reflects the
\emph{true} discriminal differences between stimuli. This
overestimation, in turn, results in spurious conclusions about these
differences \citep{McElreath_2020, Wu_et_al_2022} and, by extension,
about the underlying discriminal processes of stimuli.
Figure~\ref{fig-dispersion} also illustrates this scenario when the
model's discriminal difference distribution aligns with the thick
continuous line for \(\sigma_{B}-\sigma_{A}=0\), while the \emph{true}
discriminal difference follows any discontinuous line where
\(\sigma_{B}-\sigma_{A} \neq 0\). Furthermore, if researchers
acknowledge that misfit statistics may represent texts with different
dispersions or outlying observations, the common CJ practice of
excluding stimuli based on these statistics, as seen in
\citet{Pollitt_2012a}, \citet{Pollitt_2012b},
\citet{vanDaal_et_al_2016}, and \citet{Goossens_et_al_2018}, may
unintentionally discard valuable information, introducing bias into the
trait estimates \citep{Zimmerman_1994, McElreath_2020}. The direction
and magnitude of these biases remain unpredictable, as they depend on
which stimuli researchers exclude from the analysis.

@@

Finally, similar to what we discussed in the previous section, removing
\emph{misfit} judges--that is, judges whose assessments deviate markedly
from the shared consensus
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018, Wu_et_al_2022},
and may appear as outliers under the BTL model
\citep{Wu_et_al_2022}--risk discarding valuable information and
introducing bias into trait estimates. The direction and magnitude of
these biases remain unpredictable, as they depend on which judges
researchers exclude from the analysis
\citep{Zimmerman_1994, OHagan_2018, McElreath_2020}.

@@

Thus, to address the limitations of Thurstone's Case V and the BTL
model, this study extends Thurstone's general form using a systematic,
integrated approach that combines causal and Bayesian inference methods.
The approach begins by formulating a conceptual model represented by a
Structural Causal Model (SCM) and a Directed Acyclic Graph (DAG). This
model integrates Thurstone's core theoretical principles, such as the
discriminal processes of stimuli, alongside key CJ assessment design
features, including judges' bias, sampling procedures, and comparison
mechanisms, thereby disentangling the causal processes in the CJ system.

@@

Beyond these potential benefits, the approach offers two additional
advantages. First, it clarifies the roles and interactions of all actors
and processes involved in CJ assessments. Second, it shifts the analytic
paradigm from passively accepting the assumptions of the BTL model to
actively testing their fit with observed data. Together, these
advantages establish a principled framework for evaluating best
practices in designing CJ assessments, one that better aligns with the
demands of contemporary CJ contexts \citep{Kelly_et_al_2022}, offering
new insights into existing research and opening promising avenues for
future inquiry.

\subsection{Future research directions based on this study's
approach}\label{sec-discussion_RA}

Among the many potential directions for future research--both current
and new--three avenues deserve particular attention due to their direct
impact on the reliability and validity of trait estimates, as well as on
the accuracy of statistical inferences. The following sections outline
these avenues and explain how our approach provides a rigorous research
framework to investigate them.

@@

Although sampling and comparison mechanisms are central to modern CJ
assessments, it is striking that most of the CJ literature has examined
them only within a limited scope. Researchers have primarily
investigated the effects of adaptive comparative judgment (ACJ) designs
on trait reliability
\citep{Pollitt_2012a, Pollitt_2012b, Bramley_2015, Verhavert_et_al_2022, Mikhailiuk_et_al_2021, Gray_et_al_2024}
or proposed practical guidelines for the number of comparisons judges
should make \citep{Verhavert_et_al_2019, Crompvoets_et_al_2022}. While
these studies offer valuable insights, they often overlook the broader
role these mechanisms play within the CJ system. Moreover, as this
oversight likely stems from a broader lack of conceptual clarity about
how these mechanisms function within the system, this study formally
incorporates these mechanisms into CJ's conceptual model.

@@

However, many modern CJ applications rely on more complex assessment
designs in which the sampling and comparison mechanisms introduce more
intricate forms of missingness--such as \emph{missing at random} (MAR)
or \emph{missing not at random} (MNAR) \citep{Little_et_al_2020}. A
prominent example is the previously discussed ACJ designs, where prior
judgment outcomes inform the selection of stimulus pairs for subsequent
comparisons \citep{Pollitt_2012a, Pollitt_2012b, Bramley_2015}. This
outcome-informed pair selection suggests that the comparison mechanism
is outcome-dependent, potentially classifying ACJ as a generator of MNAR
data. This, in turn, facilitates a more nuanced understanding of the
mixed evidence surrounding ACJ's capabilities. While some studies report
that the method improves trait reliability
\citep{Pollitt_et_al_2003, Pollitt_2012a, Pollitt_2012b}, others argue
that it artificially inflates these gains
\citep{Bramley_2015, Bramley_et_al_2019, Crompvoets_et_al_2020, Crompvoets_et_al_2022},
potentially as a result of outcome-dependent missingness.

Nevertheless, regardless of the underlying missingness mechanisms, any
CJ assessment design would benefit from explicitly defining its
assumptions--a practice supported by our proposed approach. This clarity
enables researchers to evaluate how the sampling and comparison
mechanisms affect trait estimation and statistical inference within each
design. Such assessments are particularly relevant given the common
misconception in the CJ literature that Thurstone's model can naturally
handle even non-random missing data without compromising the reliability
or validity of trait estimates \citep{Bramley_2008}.

@@

Moreover, since judges draw on a variety of intricate stimulus
characteristics when evaluating complex, multidimensional traits
\citep{vanDaal_et_al_2016, Lesterhuis_et_al_2018, Chambers_et_al_2022},
and these characteristics account for variation in judgment accuracy
\citep{Gill_et_al_2013, vanDaal_et_al_2017, vanDaal_2020, Gijsen_et_al_2021},
it is reasonable to expect that judges' assessments also vary with
individual attributes such as gender, age, culture, income, education,
training, or expertise
\citep{Kelly_et_al_2022, Bartholomew_et_al_2020a}. Thus, building on the
discussion in Section~\ref{sec-discussion_RA1}, researchers could
investigate how the selection of judges influences the formation of a
``shared consensus'' and whether these individual attributes introduce
systematic biases or distortions in the observed trait distribution
\citep{Deffner_et_al_2022}. Furthermore, if such attributes indeed
undermine the assumption of ``sample-freeness,'' then it becomes
essential to explore strategies to control their influence and to
determine how many judges--and how many judgments per judge--are needed
to produce reliable trait estimates in these contexts. Additionally, it
is worth considering whether \emph{repeated measures designs}, in which
judges evaluate the same stimulus pairs multiple times
\citep{Lawson_2015}, can improve judgment consistency and accuracy.

@@

{LAST THING TO DO}

@@

Conversely, this theoretical commitment to causal inference also
introduces several practical challenges that applied CJ researchers must
navigate. These fall into two main categories: first, acquiring the
foundational knowledge necessary to apply the approach effectively; and
second, dedicating greater attention to both conceptual and statistical
modeling.

>>>>

To address these issues, this study extends Thurstone's general form
through a systematic and integrated approach that combines causal and
Bayesian inference methods. In addition to potentially enhancing
measurement reliability and validity, and improving statistical accuracy
in hypothesis testing, this approach offers two key advantages. First,
it clarifies the interactions among all actors and processes involved in
CJ assessments. Second, it shifts the current comparative data analysis
paradigm from passively accepting Case V and the BTL model assumptions
to actively testing whether those assumptions fit the data under
analysis.

@@

Significant statistical and measurement issues can arise when the
assumption of equal dispersions between stimuli does not hold.
Specifically, the BTL model may overestimate the trait's reliability,
that is, the degree to which the outcome accurately reflects the
\emph{true} discriminal differences between stimuli. This
overestimation, in turn, results in spurious conclusions about these
differences \citep{McElreath_2020, Wu_et_al_2022} and, by extension,
about the underlying discriminal processes of stimuli.
Figure~\ref{fig-dispersion} also illustrates this scenario when the
model's discriminal difference distribution aligns with the thick
continuous line for \(\sigma_{B}-\sigma_{A}=0\), while the \emph{true}
discriminal difference follows any discontinuous line where
\(\sigma_{B}-\sigma_{A} \neq 0\). Furthermore, if researchers
acknowledge that \emph{misfit statistics} may represent texts with
different dispersions or outlying observations, the common CJ practice
of excluding stimuli based on these statistics may unintentionally
discard valuable information \citep{Miller_2023}, introducing bias into
the trait estimates \citep{Zimmerman_1994, McElreath_2020}. The
direction and magnitude of these biases remain unpredictable, as they
depend on which stimuli researchers exclude from the analysis.

@@

Finally, similar to what we discussed in the previous section, removing
\emph{misfit} judges--that is, judges whose assessments deviate markedly
from the shared consensus
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018, Wu_et_al_2022},
and may appear as outliers under the BTL model
\citep{Wu_et_al_2022}--risk discarding valuable information and
introducing bias into trait estimates \citep{Miller_2023}. The direction
and magnitude of these biases remain unpredictable, as they depend on
which judges researchers exclude from the analysis
\citep{Zimmerman_1994, OHagan_2018, McElreath_2020}.

@@

Thus, to address the limitations of Thurstone's Case V and the BTL
model, this study extends Thurstone's general form using a systematic,
integrated approach that combines causal and Bayesian inference methods.
The approach begins with the development of a conceptual model,
represented as a Structural Causal Model (SCM) and a Directed Acyclic
Graph (DAG)
\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}. This
model integrates Thurstone's core theoretical principles, such as the
discriminal processes of stimuli, alongside key CJ assessment design
features, including judges' bias, sampling procedures, and comparison
mechanisms, thereby disentangling the causal processes underlying the CJ
system.

@@

Beyond these potential benefits, the approach offers two additional
advantages. First, it clarifies the roles and interactions of all actors
and processes involved in CJ assessments. Second, it shifts the analytic
paradigm from passively accepting the assumptions of Case V and the BTL
model to actively testing their fit with observed data. Together, these
advantages establish a principled framework for evaluating best
practices in designing CJ assessments, one that better aligns with the
demands of contemporary CJ contexts \citep{Kelly_et_al_2022}, offering
new insights into existing research and opening promising avenues for
future inquiry.

\subsection{Future research directions using our
approach}\label{sec-discussion_RA}

Among the many potential directions for future research--both current
and new--three avenues deserve particular attention due to their direct
impact on the reliability and validity of CJ trait estimates, as well as
on the accuracy of statistical inferences. The following sections
outline these avenues and explain how our approach facilitates their
investigation.

@@

Although sampling and comparison mechanisms are central to modern CJ
assessments, it is striking that most of the CJ literature has examined
them only within a limited scope. Researchers have primarily
investigated the effects of adaptive comparative judgment (ACJ) designs
on trait reliability
\citep{Pollitt_2012a, Pollitt_2012b, Bramley_2015, Verhavert_et_al_2022, Mikhailiuk_et_al_2021, Gray_et_al_2024}
or proposed practical guidelines for the number of comparisons judges
should make \citep{Verhavert_et_al_2019, Crompvoets_et_al_2022}. While
these studies offer valuable insights, they often overlook the broader
role these mechanisms play within the CJ system. As this oversight
likely stems from a broader lack of conceptual clarity about how these
mechanisms function within the system, this study decided to formally
incorporate these mechanisms into CJ's conceptual model.

@@

However, many modern CJ applications rely on more complex assessment
designs, in which the sampling and comparison mechanisms introduce more
intricate forms of missingness --- such as \emph{missing at random}
(MAR) or \emph{missing not at random} (MNAR) \citep{Little_et_al_2020}.
A prominent example is the previously discussed ACJ designs, where prior
judgment outcomes inform the selection of stimulus pairs for subsequent
comparisons \citep{Pollitt_2012a, Pollitt_2012b, Bramley_2015}. This
pair selection process suggests that ACJ's comparison mechanism is
outcome-dependent, potentially classifying the method as a generator of
MNAR data. If this classification holds, the mixed findings on ACJ's
capabilities become more comprehensible: some studies find that the
method improves trait reliability
\citep{Pollitt_et_al_2003, Pollitt_2012a, Pollitt_2012b}, while others
contend that it artificially inflates these gains
\citep{Bramley_2015, Bramley_et_al_2019, Crompvoets_et_al_2020, Crompvoets_et_al_2022}.

Nevertheless, regardless of the underlying missingness mechanisms, any
CJ assessment design would benefit from explicitly defining its
assumptions--a practice supported by our approach. This clarity enables
researchers to evaluate how the sampling and comparison mechanisms
affect trait estimation and statistical inference within each design.
Such assessments are particularly relevant given the common
misconception in the CJ literature that Thurstone's model can naturally
handle even non-random missing data without compromising the reliability
or validity of trait estimates \citep{Bramley_2008}.

@@

Moreover, since judges draw on a variety of intricate stimulus
characteristics when evaluating complex, multidimensional traits
\citep{vanDaal_et_al_2016, Lesterhuis_et_al_2018, Chambers_et_al_2022},
and these characteristics account for variation in judgment accuracy
\citep{Gill_et_al_2013, vanDaal_et_al_2017, vanDaal_2020, Gijsen_et_al_2021},
it is reasonable to expect that judges' assessments also vary with
attributes such as gender, age, culture, income, education, training, or
expertise \citep{Kelly_et_al_2022}. Prior studies support this view
\citep{Bartholomew_et_al_2020a, McMahon_et_al_2015}. Thus, building on
the discussion in Section~\ref{sec-discussion_RA1}, researchers could
further explore how judges' selection influences the formation of a
``shared consensus'' and whether these attributes introduce systematic
biases or distortions in the observed trait distribution
\citep{Deffner_et_al_2022}. Furthermore, if such attributes indeed
undermine the assumption of ``sample-freeness,'' researchers should
explore ways to control their influence and determine how many
judges--and how many judgments per judge--are needed to produce reliable
trait estimates in these contexts. Additionally, it is worth considering
whether \emph{repeated measures designs}, in which judges evaluate the
same stimulus pairs multiple times \citep{Lawson_2015}, can improve
judgment consistency and accuracy. Our approach provides the necessary
structure to rigorously investigate these questions.

@@

Although the CJ literature clearly defines \emph{misfit} judges and
stimuli, researchers have rarely examined how these observations relate
to Thurstonian theory. In particular, they have not identified which
elements of Thurstone's theory might explain the occurrence of misfits.
This disconnect likely stems from the fact that CJ researchers derive
misfit statistics from methods such as residual analysis and outlier
detection rather than from Thurstonian principles
\citep{Pollitt_2012a, Pollitt_2012b, Wu_et_al_2022}. They typically
define \emph{misfit judges} as those whose assessments diverge
significantly from the ``shared consensus''
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018, Wu_et_al_2022},
and \emph{misfit stimuli} as those that elicit more judgment
discrepancies than others
\citep{Pollitt_2004, Pollitt_2012a, Pollitt_2012b, Goossens_et_al_2018}.
Both definitions closely mirror the statistical concept of
outliers--that is, observations that deviate markedly from the rest of
the sample in which they occur \citep{Grubbs_1969}. However, this
resemblance extends beyond these definitions. CJ researchers also
identify misfits using standard outlier detection procedures, such as
transforming BTL model residuals into statistics of interest and
comparing them against a predefined threshold.

{After identifying misfit observations, researchers often propose
various hypotheses to explain their presence. Yet, without providing
evidence for any of these explanations, they typically exclude these
observations from analysis. This exclusion process raises three major
concerns: one related to the reliability of the methods used to identify
misfits, a second concerning the potential impact of their removal on
trait estimates and inferences, and a third regarding the hypotheses
used to explain the presence and exclusion of misfits.}

{First, regarding the reliability of identification methods, it is
relevant to note that outlying observations can only be defined relative
to a specific model \citep{McElreath_2020}. Consequently, methods based
on models such as the BTL--which rest on strong assumptions--should be
treated with caution, as these assumptions may prevent the model from
accurately capturing the underlying data-generating process of the CJ
system, thereby potentially compromising the identification of misfits.}

{Second, concerning the impact of removing misfits, the statistical
literature warns that ad-hoc outlier exclusion procedures risk
discarding valuable information \citep{Miller_2023} and introducing bias
into trait estimates. The direction and magnitude of this bias are often
unpredictable, as they depend on which observations researchers exclude
from the analysis \citep{Zimmerman_1994, OHagan_2018, McElreath_2020}.}

{Lastly, regarding the hypotheses explaining the presence and exclusion
of misfits, it is important to note that a significant number of these
hypotheses could not be tested within the BTL model due to its strong
assumptions and limited flexibility. This is the reason why there is a
lack of supporting evidence for misfit observations. In contrast, our
approach enables researchers to test these hypotheses. For instance,
researchers could examine whether misfit judges are simply those who
exhibit (an outlying degree of) systematic bias or whether they display
greater variability in their judgments compared to their peers.
Similarly, in the case of misfit stimuli, researchers could investigate
whether their status reflects an increased variability in their
discriminal processes or stems from their classification as outlying
stimuli.}

{Moreover, since outliers are not necessarily ``bad data''
\citep{McElreath_2020}, our approach offers a principled alternative
that retains misfits in the analysis without compromising trait
estimation. This alternative involves adapting our proposed models into
more sophisticated, robust measurement models
\citep{Wainer_et_al_1978}--a broad class of procedures designed to
reduce the sensitivity of parameter estimates to mild or moderate
departures from model assumptions \citep{Everitt_et_al_2010}.}

@@

Moreover, this theoretical commitment to causal inference also
introduces several practical challenges that applied CJ researchers must
navigate. These fall into two main categories: first, acquiring the
foundational knowledge necessary to apply the approach effectively; and
second, dedicating greater attention to both conceptual and statistical
modeling.
